---
title: "Getting Started with peruopen"
format: html
vignette: >
  %\VignetteIndexEntry{Getting Started with peruopen}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

```{r}
#| label: setup
library(peruopen)
```

## Overview

The **peruopen** package provides a comprehensive interface to Peru's CKAN data portal ([datosabiertos.gob.pe](https://datosabiertos.gob.pe)), enabling you to search, discover, and download open government datasets directly into R.

With **peruopen**, you can access over 3,954 datasets and 10,000+ resources from 283+ government organizations in Peru.

## Installation

Install the package from GitHub:

```{r, eval=FALSE}
# Install from GitHub
devtools::install_github("horaciochacon/peruopen")
library(peruopen)
```

## Basic Usage

### 1. Get the Complete Data Catalog

Start by getting the complete catalog of all available datasets:

```{r}
# Get the complete catalog (may take a few minutes on first run)
catalog <- po_catalog(refresh = TRUE)

# Explore the catalog structure
catalog$datasets
catalog$resources
```

The catalog contains:
- **datasets**: Information about all available datasets
- **resources**: Details about individual files and data resources  
- **summary**: Statistics about the catalog

### 2. Search for CDC Data

Search for surveillance data from CDC (Centro Nacional de Epidemiología):

```{r}
# Search for datasets with CDC tags
vigilancia <- po_search(tags = "cdc")
vigilancia

# Explore the results
vigilancia$datasets
vigilancia$resources
```

### 3. Download CDC Data

Download resources directly into R:

```{r}
# Download all resources from the CDC search
vigilancia <- po_get(vigilancia$resources)

names(vigilancia)
```

### 4. Search by Organization

Find data from specific government organizations like MINSA:

```{r}
# Search for all MINSA (Ministry of Health) datasets
minsa <- po_search(organizations = "MINSA")
minsa
minsa$datasets
```

### 5. Geographic Data from INEI

Search for ubigeo (geographic codes) data from INEI:

```{r}
# Search for ubigeo data from INEI using both tags and organization
ubigeo <- po_search(tags = "ubigeo", organizations = "INEI")
ubigeo
ubigeo$datasets
ubigeo$resources

# Download the ubigeo data
po_get(ubigeo$resources[4,])
```

### 6. Survey Data Analysis

Search for demographic survey data:

```{r}
# Search for ENDES (Demographic and Health Survey) data from INEI
endes <- po_search(query = "ENDES", organizations = "INEI")
endes
endes$datasets
```

## Advanced Search Features

### Format-specific Downloads

```{r}
# Find all CSV files about education
education_csv <- po_search("educacion", formats = "CSV")
education_csv$resources

# Get only Excel files
excel_data <- po_search("educacion", formats = c("XLSX", "XLS"))
excel_data$resources
```

### Combined Search Criteria

```{r}
# Find health datasets from MINSA in CSV format
health_csv <- po_search(
  query = "salud",
  organizations = "MINSA", 
  formats = "CSV"
)
health_csv

# Search by multiple tags
health_tags <- po_search(tags = c("salud", "epidemiologia"))
health_tags
```

### Save Files Locally

```{r, eval=FALSE}
# Download and save files to local directory
po_get("dataset-name", save_to = "data/peru/")
```

## Common Use Cases

### Health Data Analysis

```{r}
# Get all health datasets from MINSA
po_search(organizations = "MINSA")

# Find COVID-19 related data
po_search("covid")

# Get malaria surveillance data
 po_search("malaria", formats = "CSV")
```

### Economic and Budget Data

```{r}
# Find budget execution data from MEF
po_search(organizations = "MEF")

# Search for economic indicators
po_search("economico")
```


## Data Coverage

Peru's open data portal includes datasets from:

- **MINSA** - Ministry of Health (epidemiological data, health statistics)
- **MEF** - Ministry of Economy (budget execution, economic indicators)
- **MINEDU** - Ministry of Education (educational statistics)
- **INEI** - National Statistics Institute (census, demographic data)
- **Regional Governments** - Local statistics and programs
- **283+ organizations** total

The portal contains approximately 1.7 TB of data across all resources.

## Getting Help

- Use `?function_name` for help with specific functions
- Browse the [function reference](https://horaciochacon.github.io/peruopen/reference/)
- Report issues at [GitHub](https://github.com/horaciochacon/peruopen/issues)