---
title: "Testing Quarto Output for peruopen"
format: html
editor: visual
---

# Testing knit_print Methods

This document tests the new `knit_print` methods for the peruopen package to ensure that output renders nicely in Quarto/markdown environments.

```{r setup}
#| message: false
library(peruopen)
```

## Test po_search Output

```{r search-test}
# Test search results
results <- po_search("salud", max_results = 10)
results
```

## Test po_explore Output

```{r explore-test}
# Test exploration results  
exploration <- po_explore("educacion", max_datasets = 5)
exploration
```

## Test po_datasets Output

```{r datasets-test}
# Test datasets display
datasets <- po_list_datasets(limit = 5)
datasets
```

## Test po_resources Output

```{r resources-test}
# Test resources display
resources <- po_list_resources(limit = 5)
resources
```

## Console vs. Quarto Comparison

The above outputs should render as nicely formatted markdown with:

- Headers (`#` and `##`)
- Bold text (`**bold**`)
- Italic text (`*italic*`)
- Bullet points (`-`)
- Horizontal rules (`---`)
- Code formatting (`` `code` ``)

While in the console, the same objects should display with CLI colors and formatting.